---
title:  "Assignment_2"
author: "Kris Evers"
date:   "2019-09-23"
output: html_notebook
---

## Scientific Programming | Assignment 2

In this assignment information about the boiling point of simple molecules is requested by running a SPARQL query on the Wikidata database. Alkanes are molecules with only single bonds. It is straightforward to see a relationship between the number of carbon atoms a molecule has and the boiling point. The determination is more complicated than that though. Molecules with the same number of atoms can have different structures. A branched alkane has a different boiling point than a unbranched alkane with the same number of carbon atoms. In this assignment the Partial Least Squares method is used to determine what the best predictor(s) is/are for the measure of boiling point.

### Setup of dependencies

```{r library, echo = FALSE, error=FALSE}
  # install packages
  install.packages(c("wikidataQueryServiceR", "rJava", "rcdk"))
```

### SPARQL

```{r query}
  # load wikidataQueryServiceR package
  library(WikidataQueryServiceR)

  # define and request query
  r <- query_wikidata('SELECT ?ane ?aneLabel ?bp ?bpUnit ?bpUnitLabel ?CC 
                      WHERE {   ?ane wdt:P31/wdt:P279* wd:Q41581 ;   
                      wdt:P233 ?CC ;     
                      p:P2102 [         ps:P2102 ?bp ;           
                      psv:P2102/wikibase:quantityUnit  ?bpUnit         ] .   
                      SERVICE wikibase:label { bd:serviceParam wikibase:language "[AUTO_LANGUAGE],en". } } 
                      ')
```

```{r bp conversion}
  # conversion of boiling point temperatures to degree kelvin
  r$bp[r$bpUnitLabel == "degree Celsius"] <- r$bp[r$bpUnitLabel == "degree Celsius"] + 273.15
  r$bpUnitLabel[r$bpUnitLabel == "degree Celsius"] <- "kelvin"
  r$bp[r$bpUnitLabel == "degree Fahrenheit"] <- (r$bp[r$bpUnitLabel == "degree Fahrenheit"]-32)*5/9 + 273.15
  r$bpUnitLabel[r$bpUnitLabel == "degree Fahrenheit"] <- "kelvin"
```

```{r smiles}
  library(rJava)
  library(rcdk)

  parsed_smiles <- parse.smiles(r$CC)

  descNames <- c(
  'org.openscience.cdk.qsar.descriptors.molecular.FragmentComplexityDescriptor',
  'org.openscience.cdk.qsar.descriptors.molecular.WienerNumbersDescriptor',
  'org.openscience.cdk.qsar.descriptors.molecular.FMFDescriptor'
  )

  descs <- eval.desc(parsed_smiles, descNames)
```

```{r pls}
  
```
